% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.

% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
%\renewcommand\UrlFont{\color{blue}\rmfamily}

\usepackage{amsmath}
\DeclareMathOperator{\DOA}{DOA}
\DeclareMathOperator{\FA}{FA}
\DeclareMathOperator{\DL}{DL}
\DeclareMathOperator{\AC}{AC}

\usepackage{xspace}
\newcommand{\dplyr}{\texttt{dplyr}\xspace}
\newcommand{\m}{\texttt}

\begin{document}
%
\title{Using Julia-Embedded Datalog for Data Analysis}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{%
First Author\inst{1}%\orcidID{0000-1111-2222-3333}
\and
Second Author\inst{1}%\orcidID{1111-2222-3333-4444}
}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Northeastern University}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The current approach to building data analysis applications is
dominated by languages like Python, R, and, to some extent, Julia:
high-level imperative languages with thorough support for the typical
data processing tasks by the way of libraries. Imperative features like
mutation may hinder understanding of a script and its validity, while
libraries pose the challenge of learning new APIs and track changes in
the old ones. Both of these issues can be resolved by the employment of
a declarative domain-specific language built for data manipulation.
As a bonus, a significantly special-purposed DSL shall
allow efficient compilation to a robust application~--- something that
has been an issue with languages like Python and R for decades. The
only issue with a declarative interface would be the lack of support
for inherently imperative tasks (e.g. file system manipulation), which,
we argue, can be resolved by embedding the DSL into a general-purpose
language.

In this paper, we showcase a commercial implementation of Datalog
embedded in Julia, called Delve, that checks all the above boxes: it is declarative,
has virtually zero amount of API vocabulary needed to apply it,
and can resort to the efficient JIT compiler of Julia for the tasks
not amenable for declarative processing. We consider two implementations
of a data processing application, using Delve and R, and compare them
along various axes, most notably linguistics and performance.


%The abstract should briefly summarize the contents of the paper in
%150--250 words.

\keywords{Datalog  \and Julia \and R \and embedded languages.}
\end{abstract}
%
%
%

\input{text}

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{refs.bib}
%
\end{document}
